---
title: "PA1_template"
author: "Venkataraman TK"
date: "14/04/2020"
output: html_document
---
# Reproducible Research:  Assessment 1

## Reading the dataset and viewing the dataset summary
```{r loaddata}
unzip(zipfile="activity.zip")
df <- read.csv("activity.csv")
summary(df)
```

## Histogram Plot of total number of Steps taken each day, including Mean and Median steps taken each day
```{r}
library(ggplot2)
total.steps <- tapply(df$steps, df$date, FUN=sum, na.rm=TRUE)
qplot(total.steps, binwidth=1000, xlab="Total number of steps taken each day")
mean(total.steps, na.rm=TRUE)
median(total.steps, na.rm=TRUE)
```

## Time series plot of the average number of steps taken
```{r}
library(ggplot2)
averages <- aggregate(x=list(steps=df$steps), by=list(interval=df$interval),
                      FUN=mean, na.rm=TRUE)
ggplot(data=averages, aes(x=interval, y=steps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken")
```

## Identifying the 5-minute interval that, on average contains the max number of steps
```{r}
averages[which.max(averages$steps),]
```

## Imputing the missing values
From the dataset summary, there were many missing values

```{r how_many_missing}
missing <- is.na(df$steps)
# How many missing
table(missing)
```

To impute the missing values, they are filled in with mean value for that 5-minute interval.

```{r}
# Replace each missing value with the mean value of its 5-minute interval
fill.value <- function(steps, interval) {
    filled <- NA
    if (!is.na(steps))
        filled <- c(steps)
    else
        filled <- (averages[averages$interval==interval, "steps"])
    return(filled)
}
filled.data <- df
filled.data$steps <- mapply(fill.value, filled.data$steps, filled.data$interval)
```
Now, using this data set, plotting a histogram of the total number of steps taken each day and calculate the mean and median total number of steps, gives a better understanding.

```{r}
total.steps <- tapply(filled.data$steps, filled.data$date, FUN=sum)
qplot(total.steps, binwidth=1000, xlab="total number of steps taken each day")
mean(total.steps)
median(total.steps)
```

Mean and median values are higher after imputing missing data. The reason being
the `NA` values. However, after replacing missing `steps` values with the mean `steps`
of associated `interval` value, these 0 values are removed from the histogram
of total number of steps taken each day.

## Are there differences in activity patterns between weekdays and weekends?

Identifying the Days of the week to utilize in the panel plot generation.

```{r}
weekday.or.weekend <- function(date) {
    day <- weekdays(date)
    if (day %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"))
        return("weekday")
    else if (day %in% c("Saturday", "Sunday"))
        return("weekend")
    else
        stop("invalid date")
}
filled.data$date <- as.Date(filled.data$date)
filled.data$day <- sapply(filled.data$date, FUN=weekday.or.weekend)
```

Generating a panel plot containing plots of average number of steps taken
on weekdays and weekends.
```{r}
averages <- aggregate(steps ~ interval + day, data=filled.data, mean)
ggplot(averages, aes(interval, steps)) + geom_line() + facet_grid(day ~ .) +
    xlab("5-minute interval") + ylab("Number of steps")
```